Short-time simulations of open collective quantum spin systems

### 2019-04-20 soc_squeezing/collective_paper/collective_spin.tex
added OTOC figure and commentary

\begin{abstract}
  We present a new method to compute short-time expectation values in
  large ensembles of collective spin systems with generic Markovian
  decoherence.  Our method is based on a Taylor expansion of a formal
  solution to the equations of motion for Heisenberg operators.  This
  expansion can be truncated at finite order to obtain virtually exact
  results at short times that are relevant for metrological
  applications such as spin squeezing.  In order to evaluate the
  expansion for Heisenberg operators, we compute the relevant
  structure constants of a collective operator algebra.  We
  demonstrate the utility of our method by computing spin squeezing
  and two-time correlation functions in strong-decoherence regimes
  that are inaccessible to existing numerical methods.  Our method can
  be straightforwardly generalized to the case of a collective spin
  coupled to bosonic modes, relevant for trapped ion and cavity QED
  experiments, and may be used to compute quantities such as
  out-of-time-ordered correlators that can feature short-time
  signatures of quantum chaos and information scrambling.
\end{abstract}

@@

Second, the TST expansion enables computing spin correlators in
strong-decoherence regimes entirely inaccessible to other methods.  As
an example, Figure \ref{fig:decoherence_strong} shows squeezing of
$N=10^4$ spins initially in $\ket{\X}$, undergoing spontaneous decay,
excitation, and dephasing of individual spins at rates
$\gamma_-=\gamma_+=\gamma_\z=100\chi$.  These results show that TNT is
much more robust to this kind of decoherence than OAT or TAT, a
finding that could not be deduced from the weak-decoherence
simulations presented in Figure \ref{fig:benchmarking}.
Strong-decoherence computations of this sort were used to put lower
bounds on theoretically achievable spin squeezing via TAT with
decoherence in Ref.~[\citenum{he2019engineering}], exemplifying a
concrete application of the TST expansion and the collective-spin
structure constants calculated in this work.

\begin{figure*}
  \centering
  \includegraphics{two_time.pdf}
  \caption{Magnitude and phase of the two-time connected correlator
    $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\Gamma_\a$ or $\gamma_\a$, respectively
    corresponding to collective or single-spin decoherence through
    jump operators $\hat S_\a$ or $\hat s_\a^{(j)}$.  Collective
    decoherence rates ($\Gamma_\a$) were chosen to be comparable to
    the OAT squeezing strength $\chi$, while single-spin decoherence
    rates ($\gamma_\a$) were scaled by an additional factor of $S=N/2$
    in order to yield comparable effects on the correlator $C\p{t}$ at
    the times shown.  At short times, the correlations captured by
    $C\p{t}$ are essentially insensitive to collective dephasing
    ($\Gamma_\z$), while the magnitude $\abs{C\p{t}}$ is generally
    suppressed by other forms of decoherence.  Whether decoherence
    occurs through spin decay, excitation, or dephasing also leaves a
    signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results computed
    using the TST expansion in Eq.~\eqref{eq:TST} with $M=35$.}
  \label{fig:two_time}
\end{figure*}

\section{Two-time correlation functions}
\label{sec:two_time}

As a final example of collective-spin physics that is numerically
accessible via the TST expansion of Heisenberg operators, we consider
the calculation of collective-spin two-time correlation functions.  In
particular, we consider the effect of decoherence on short-time
behavior of the two-time connected correlator
\begin{align}
  C\p{t}
  \equiv \f1S\sp{\bk{\hat S_+\p{t} \hat S_-\p{0}}
    - \bk{\hat S_+\p{t}}\bk{\hat S_-\p{0}}}
  \label{eq:two_time}
\end{align}
in the context of the squeezing models in Section \ref{sec:squeezing}.
In an equilibrium setting, correlation functions of this sort contain
information about the linear response of Heisenberg operators to
perturbations of a system; in a non-equilibrium setting, they
contribute to short-time linear response (see Appendix
\ref{sec:linear_response}).  Similar correlators have made appearances
as order parameters that diagnose time-crystalline phases of
matter\cite{tucker2018shattered}, and related multi-time correlators,
known as out-of-time-ordered correlators, are commonly examined for
signatures of quantum chaos and information
scrambling\cite{maldacena2016bound}.

Figure \ref{fig:two_time} shows the behavior of $C\p{t}$ for $N=10^4$
spins, initially in the state
$\ket\X\propto\p{\ket\up+\ket\dn}^{\otimes N}$, evolving under the
squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} both
with and without various sources of decoherence.  The two-time
correlations captured by $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$
exhibit signatures of the type of decoherence present, with e.g.~clear
dependence of the phase $\phi\p{t}$ on spin decay or excitation.
Figure \ref{fig:two_time} serves as an example for the type of
behavior that is accessible at short times with the TST expansion.
This example is straightforward to extend to multi-time and
out-of-time-ordered correlators, as well as equilibrium settings.

@@

We have presented a new method for computing correlators in collective
spin systems at short times.  This method is based on truncating a
short-time expansion of Heisenberg operators, and can access
correlators on time scales that are relevant to metrological
applications such as spin squeezing.  In order to evaluate the
truncated short-time (TST) expansion of Heisenberg operators, we have
computed the structure constants of a collective spin operator
algebra, which we hope will empower future analytical and numerical
studies of collective spin systems.  Even though we considered only
non-equilibrium spin-squeezing processes in this work, our method can
be applied directly in an equilibrium setting, and is straightforward
to generalize to systems such as trapped ions and optical cavities
with collective spin-boson interactions.  In such contexts, our method
may be used to benchmark the short-time effects of decoherence, or
compute quantities such as out-of-time-ordered correlators that
diagnose the onset of quantum chaos and information scrambling.

>>>>

\begin{abstract}
  We present a new method to compute short-time expectation values in
  large collective spin systems with generic Markovian decoherence.
  Our method is based on a Taylor expansion of a formal solution to
  the equations of motion for Heisenberg operators.  This expansion
  can be truncated at finite order to obtain virtually exact results
  at short times that are relevant for metrological applications such
  as spin squeezing.  In order to evaluate the expansion for
  Heisenberg operators, we compute the relevant structure constants of
  a collective operator algebra.  We demonstrate the utility of our
  method by computing spin squeezing, two-time correlation functions,
  and out-of-time-ordered correlators for $10^4$ spins in
  strong-decoherence regimes that are otherwise inaccessible via
  existing numerical methods.  Our method can be straightforwardly
  generalized to the case of a collective spin coupled to bosonic
  modes, relevant for trapped ion and cavity QED experiments, and may
  be used to investigate short-time signatures of quantum chaos and
  information scrambling.
\end{abstract}

@@

Second, the TST expansion enables computing spin correlators in
strong-decoherence regimes entirely inaccessible to other methods.  As
an example, Figure \ref{fig:decoherence_strong} shows squeezing of
$N=10^4$ spins initially in $\ket{\X}$, undergoing spontaneous decay,
excitation, and dephasing of individual spins at rates
$\gamma_-=\gamma_+=\gamma_\z=100\chi$.  These results show that TNT is
much more robust to this kind of decoherence than OAT or TAT, a
finding that could not be deduced from the weak-decoherence
simulations presented in Figure \ref{fig:benchmarking}.
Strong-decoherence computations of this sort were used to put lower
bounds on theoretically achievable spin squeezing via TAT with
decoherence in Ref.~[\citenum{he2019engineering}], exemplifying a
concrete and practical application of the TST expansion and the
collective-spin structure constants calculated in this work.

\section{Two-time correlation functions and out-of-time-ordered
  correlators}
\label{sec:multi_time}

\begin{figure*}
  \centering
  \subfloat[Two-time correlator
  $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
  Eq.~\eqref{eq:two_time}. \label{fig:two_time}]
  {\includegraphics{two_time.pdf}} \\
  \subfloat[Squared commutator $D\p{t}$ defined in
  Eq.~\eqref{eq:OTOC}. \label{fig:OTOC}]
  {\includegraphics{OTOC.pdf}}
  \caption{The two-time connected correlator $C\p{t}$ and squared
    commutator $D\p{t}$, respectively defined in
    Eqs.~\eqref{eq:two_time} and \eqref{eq:OTOC}, for $N=10^4$ spins
    initially in the polarized state
    $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$ evolving under
    the squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.
    Results are shown for time evolution without decoherence (leftmost
    panels), as well as with decoherence at rates $\Gamma_\a$ or
    $\gamma_\a$, respectively corresponding to collective or
    single-spin decoherence through jump operators $\hat S_\a$ or
    $\set{\hat s_\a^{(j)}}$.  Collective decoherence rates
    ($\Gamma_\a$) were chosen to be comparable to the OAT squeezing
    strength $\chi$, while single-spin decoherence rates ($\gamma_\a$)
    were scaled by an additional factor of $S=N/2$ in order to yield
    comparable effects on the correlators $C\p{t}$ and $D\p{t}$ at the
    times shown.  At short times, the multi-time correlations captured
    by $C\p{t}$ and $D\p{t}$ are essentially insensitive to collective
    dephasing ($\Gamma_\z$), while their magnitudes are generally
    suppressed by other forms of decoherence.  Whether decoherence
    occurs through spin decay, excitation, or dephasing also leaves a
    signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results computed
    using the TST expansion in Eq.~\eqref{eq:TST} with $M=20$.}
  \label{fig:multi_time}
\end{figure*}

As a final example of collective-spin physics that is numerically
accessible via the TST expansion of Heisenberg operators, we consider
the calculation of two-time correlation functions and
out-of-time-ordered correlators (OTOCs).  In particular, we consider
the effect of decoherence on short-time behavior of the two-time
connected correlator
\begin{align}
  C\p{t}
  \equiv \f1S\p{\bk{\hat S_+\p{t} \hat S_-\p{0}}
    - \bk{\hat S_+\p{t}}\bk{\hat S_-\p{0}}},
  \label{eq:two_time}
\end{align}
and the expectation value of a squared commutator,
\begin{align}
  D\p{t}
  \equiv \f1{S^2}\Braket{\sp{\hat S_+\p{t},\hat S_-\p{0}}_-^\dag
    \sp{\hat S_+\p{t},\hat S_-\p{0}}_-},
  \label{eq:OTOC}
\end{align}
in the context of the squeezing models in Section \ref{sec:squeezing}.
In an equilibrium setting, correlation functions similar to that in
Eq.~\eqref{eq:two_time} contain information about the linear response
of Heisenberg operators to perturbations of a system; in a
non-equilibrium setting, they contribute to short-time linear response
(see Appendix \ref{sec:linear_response}).  Similar correlators have
made appearances as order parameters that diagnose time-crystalline
phases of matter\cite{tucker2018shattered}.  OTOCs such as that in
Eq.~\eqref{eq:OTOC}, meanwhile, are commonly examined for signatures
of quantum chaos and information scrambling\cite{maldacena2016bound,
  swingle2018unscrambling, garcia-mata2018chaos}.

Figure \ref{fig:multi_time} shows the behavior of $C\p{t}$ and
$D\p{t}$ for $N=10^4$ spins, initially in the state
$\ket\X\propto\p{\ket\up+\ket\dn}^{\otimes N}$, evolving under the
squeezing Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} both
with and without various sources of decoherence.  At short times, both
correlators are insensitive to collective dephasing, while their
magnitudes are suppressed by other types of decoherence.  The
correlations captured by $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$
also exhibit signatures of the type of decoherence present, with
e.g.~clear dependence of the phase $\phi\p{t}$ on spin decay or
excitation.  Figure \ref{fig:multi_time} serves as an example for the
type of behavior that is accessible at short times with the TST
expansion.  These examples are straightforward extend to equilibrium
settings and spin-boson systems.

@@

We have presented a new method for computing correlators at short
times in collective spin systems.  This method is based on truncating
a short-time expansion of Heisenberg operators, and can access
correlators on time scales that are relevant to metrological
applications such as spin squeezing.  In order to evaluate the
truncated short-time (TST) expansion of Heisenberg operators, we have
computed the structure constants of a collective spin operator
algebra, which we hope will empower future analytical and numerical
studies of collective spin systems.  Even though we considered only
non-equilibrium spin-squeezing processes in this work, our method can
be applied directly in an equilibrium setting, and is straightforward
to generalize to systems such as trapped ions and optical cavities
with collective spin-boson interactions.  In such contexts, our method
may be used to benchmark the short-time effects of decoherence, or
study the onset of quantum chaos and information scrambling.
